---
title: "US Sex Education/Teen Pregnancy-Birth Rates - Geospatial project"
author: "Melanie Fisher"
date: "2024-05-01"
output: html_document
---
#Does a teenager's location in the United States increase ones chances of getting pregnant?

#Do counties with high teenage birthrates neighbor other counties with high teenage birthrates?
#Do states level of sexual education impact Teenage Birth Rates?
Is there a difference between rural and urban counties?
regression - can predict teenage birth rates based on sexedu

Descriptive/Exploratory - without trying to model it
spatial autocorrelation between of birthrates, if there are clusters try to make a hypothesis about why, could map it out

is religious beliefs, general school policies - clustering - local Moran I
why there are these geospatial difference


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Relevant Definitions
Pregnancy rate: refers to the rate of successful pregnancies.
Birth Rate: number of live births per thousand of population per year
```{r}
library(terra) |> suppressMessages()
library(tmap)|> suppressMessages()
library(stringr) |> suppressMessages()
library(spdep) # package for spatial dependence
library(sf)|> suppressMessages()
library(dplyr)|> suppressMessages()
library(tidyverse)|> suppressMessages()
library(maps)|> suppressMessages()
library(readxl)|> suppressMessages()
library(tigris)
library(tidycensus)
options(tigris_use_cache = TRUE)
```
#Coding
States have different policies regarding sexual education.
Some do not mandate sex education at all. Though while some do not mandate it, they still have directives about what needs to be included if the school decides to teach it anyways. I coded these as not having sexual education, to try to create a comprehensive code and level of sexual education in order to display them together.
```{r}
# Load state and county data
#states_sf <- states(class = "sf")**
#counties_sf <- counties(class = "sf")**

states <- tigris::states(cb = TRUE, class = "sf")  
counties <- tigris::counties(cb = TRUE, class = "sf")
# cb = TRUE for a low-resolution version for faster processing
# Visualize the map with state and county boundaries
#map <- mapview::mapview(states) + mapview::mapview(counties)**
#print(map)**

all_sf <- bind_rows(
  mutate(states, type = "state"),
  mutate(counties, type = "county")
)
mapview::mapview(all_sf)
```
```{r}
# Filter out US territories from dataset
# List of US states
us_states <- c("alabama", "arizona", "arkansas", "california", 
               "colorado", "connecticut", "delaware", "florida", "georgia", 
               "idaho", "illinois", "indiana", "iowa", "kansas", 
               "kentucky", "louisiana", "maine", "maryland", "massachusetts", 
               "michigan", "minnesota", "mississippi", "missouri", "montana", 
               "nebraska", "nevada", "new hampshire", "new jersey", "new mexico", 
               "new york", "north carolina", "north dakota", "ohio", 
               "oklahoma", "oregon", "pennsylvania", "rhode island", "south carolina", 
               "south dakota", "tennessee", "texas", "utah", "vermont", 
               "virginia", "washington", "west virginia", "wisconsin", "wyoming")
all_sf <- all_sf %>%
  filter(tolower(STATE_NAME) %in% us_states)
unique(all_sf$STATE_NAME)
```

```{r}
#Load birth rate data from NCHS
birthrate <- read_excel("C:/Users/melon/OneDrive/Desktop/Masters/Hertie/Geospatial/NCHS2016-Teen_Birth_Rates.xlsx")

#Load Sex Education data by State from Guttmacher Data Center
sexedu <-read_excel("C:/Users/melon/OneDrive/Desktop/Masters/Hertie/Geospatial/GuttmacherDataCenter-SexEdbyState_treated.xlsx")
```

```{r}
# Merge birthrate data with all_sf
all_sf <- all_sf %>%
  left_join(birthrate, by = c("NAME" = "County", "STATE_NAME" = "State"))

# Merge sex education data with all_sf
all_sf <- all_sf %>%
  left_join(sexedu, by = c("STATE_NAME" = "U.S. State"))
``` 


```{r}
#Determine best # of groups to put separate birthrate data into
library(cluster)  # For clustering algorithms
print(head(birthrate))
# Assuming 'birthrate' is your dataframe and 'Birth_Rate' is the column
data <- birthrate$`Birth Rate` %>% na.omit()  # Remove NA values for clustering
wss <- map_dbl(1:10, function(k) {
  kmeans(data, centers = k, nstart = 25)$tot.withinss
})

# Plotting the within-group sum of squares by number of clusters
qplot(1:10, wss, geom = "line") +
  labs(x = "Number of Clusters", y = "Total Within Sum of Squares", title = "Elbow Method for Optimal k")

```



```{r}
# Map birth rate by county

custom_palette <- colorRampPalette(c("blue", "pink", "red"))(100)  # More control over color transition

tmap_mode("plot")

tm <- tm_shape(all_sf_clean) +
  tm_polygons("Birth Rate", id = "NAME",
              palette = custom_palette, title = "Teen Birth Rate",
              style = "quantile", n = 3,
              border.col = "black", border.alpha = 0.5) +
 tm_layout(
    frame = FALSE,
    font.size = 10,  # Adjust base font size
    outer.margins = c(0, 0.02, 0.02, 0.02),  # Reduce outer margins
    legend.position = c("right", "bottom"),
    legend.bg.color = "white",  # Adding background color to legend for better readability
    legend.bg.alpha = 0.7,  # Semi-transparent background for legend
    legend.text.size = 0.8,  # Adjust legend text size
    legend.title.size = 0.9  # Adjust legend title size
  ) +
  tm_scale_bar(
    position = c("left", "bottom"),
    breaks = c(0, 100, 200, 300, 400, 500)
  )


# Print the map
print(tm)
```
```{r} 
#** MAYBE DELETE - but this is title solution
# Convert the tmap object to a grob (graphical object) for use with gridExtra
map_grob <- tmap_grob(tm)

# Create a text grob for the title
title_grob <- textGrob("Teenage Birth Rate by County", gp=gpar(fontface="bold", fontsize=20), hjust=0.5, vjust=-1)

# Arrange the title and the map using grid.arrange
grid.arrange(title_grob, map_grob, heights=c(1, 10), ncol=1)
Explanation
```



##Average Teen Birth Rate by State
```{r}
# Aggregate birth rate data to the state level by calculating the mean birth rate per state
state_birthrate <- all_sf %>%
  group_by(STATE_NAME) %>%
  summarise(Avg_Birth_Rate = mean(`Birth Rate`, na.rm = TRUE))  

states <- st_join(states, state_birthrate, join = st_intersects)
states <- states %>%
 filter(!is.na(Avg_Birth_Rate))


tm_states_birthrate <- tm_shape(states) +
  tm_polygons("Avg_Birth_Rate", id = "NAME",
              palette = colorRampPalette(c("blue", "pink", "red"))(100), 
              title = "Average Teen Birth Rate by State",
              style = "quantile", n = 3,
              border.col = "black", border.alpha = 0.5) +
  tm_layout(
    frame = FALSE,
    font.size = 10,
    outer.margins = c(0, 0.02, 0.02, 0.02),
    legend.position = c("right", "bottom"),
    legend.bg.color = "white",
    legend.bg.alpha = 0.7,
    legend.text.size = 0.8,
    legend.title.size = 0.9
  ) +
  tm_scale_bar(
    position = c("left", "bottom"),
    breaks = c(0, 100, 200, 300, 400, 500)
  )

# Print the map
print(tm_states_birthrate)
```



```{r}
#See neighbors
nb<-poly2nb(all_sf,queen=TRUE)
nb
```

```{r}
plot(all_sf$geometry, border = "lightgray")
plot.nb(nb, st_geometry(all_sf), add = TRUE)
```

```{r}
# Need to remove NAs so I can run Spatial Autocorrelations/Moran's Test
all_sf_clean <- all_sf %>%
  filter(!is.na(`Birth Rate`))
```


```{r}
# Create weights for neighbors w/ style Binary 
nb_clean <- poly2nb(all_sf_clean)

nbw <- nb2listw(nb_clean, style = "W", zero.policy = T)

#hypothesis set to "greater", meaning I expect a positive autocorrelation
gmoran <- moran.test(all_sf_clean$`Birth Rate`, nbw,
                     alternative = "greater")
gmoran
```

```{r}
moran.plot(data$ai0209, nbw, labels=F)
```


